---
title: "SIRTD Data Generating Processes"
author: "Breck Baldwin"
date:  '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA, include = TRUE)
```

# Introduction

This page contains data generating programs for a common class of epidemiological models derived from compartments for the states (S) susceptible, (I) infectious and (R) for recovered --SIR models. This page/code adds compartments for (T) terminal and (D) dead counts useful for the grim task of distinguishing the dead from recovered. There is additionally data generated for tweets but that exists in parallel with the core SIRTD model which have to sum to the population at each step so do not particpate in those counts. 

For more explanation about SIR models see [https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html](https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html) which I found to be a good explanation as well as an excellent introduction to Bayesian approaches using Stan. This is the model I used to fit this data, there is a model reproduction check list at [https://codatmo.github.io/Simple_SIR/](https://codatmo.github.io/Simple_SIR/). 

# Navigation

This Rmarkdown page exists as part of the [CoDatMo](https://codatmo.github.io/) project. The repo containing this page is at [https://github.com/codatmo/dataGeneratingProcess1/](https://github.com/codatmo/dataGeneratingProcess/) as 'SirtdGeneratingProcess.Rmd' and rendered as 'SirtdGeneratingProcess.html'. 

# Model 

The model simulates a fixed population at the individual level on a daily basis. Larger or smaller uniform units could be used. At a pseudo code level the algorithm is:

```
nPop = 1000 # population size
nPatientZero = 10 # how many start infected
nDailyContacts = 10 # number of daily contacts per day for infected
betaDailyInfectionRates[nDays] = list of infection rate for each day--can vary
gammaResolvedDailyRate = rate moving from 'i' to 'r' or 't'
dailyTweetRateInfected = rate of tweets per person that is 'i'
meanDaysToDeath = mean of how many days does it take from 't' to 'd'
Create dayState[nPop] list of all people with state = 's' for susceptible

Patient zeros: set nPatientZero number of dayState list to 'i' for infected


tweets[nDay] = count of tweets per day


for (day in numberOfDays) {
  Copy dayState to nextDayState 
  for (person in 1 to nPop) { # infections
    if (dayState[person] equals 'i') { # infecting others
      for (otherPerson in sample(numberDailyContacts, nPop))
        chanceInfection = betaDailyInfectionRates[day]/numberDailyContacts
          if (randomChance(person, chanceInfection) == infected and 
              dayState[otherPerson] equals 's') {
             nextDayState[otherPerson] = 'i' 
          }
    }
    if (dayState[person] equals 'i') { # recovery or death
      if (randomChance(person, gammaResolvedDailyRate) == resolved) {
         if (randomChoice(person, deathRate) == died) {
            nextDayState[person] = 't' #terminally ill 
         }
         else {
            nextDayState[person] = 'r' # recovered, will not die
         }
      }
    }
    if (dayState[person] equals 'i') { # do they tweet while infected
      if (randomChance(person, tweetDailyRateInfected) == tweets) {
         tweetCountForDay[day] = tweetCountForDay[day] + 1
      }
    }
    if (dayState[person] equals 't') { # does a 't' -> 'd'
      if (randomChance(person, meanDaysToDeath) == death) {
          nextDayState[person] = 'd' 
      } 
    }
   dayState = nextDayState # overwrite the current day with the day we just created
}
```
The `randomChance()` function stands in place of many functions like the Bernoulli distribution. But there are many candidates, the source below details the actual choices. 

Actual code below


```{r echo=TRUE}
SirtdVaryBeta = function(runName = runName,
                         nPop = nPop,
                         nDays = nDays,
                         print = PRINT,
                         betaDailyInfectionRates = betaDailyInfectionRates,
                         gammaResolvedPerDayRate = gammaResolvedPerDayRate,
                         deathRate = deathRate,
                         tweetRateInfected = tweetRateInfected,
                         nPatientZero = nPatientZero,
                         meanDaysToDeathFromT = meanDaysToDeathFromT,
                         nDailyContacts = nDailyContacts) {
  
  dayState = c(rep('i',nPatientZero), rep('s', nPop - nPatientZero))
  tweets = rep(0, nPop)
  colNames = c('runName', 'day', 's', 'i', 'r', 't', 'd_today', 'd', 'tweets')
  df = data.frame(matrix(nrow = 0, ncol=length(colNames)))
  colnames(df) = colNames
  nextDayState = dayState
  for (day in 1:nDays) {
    df[day,] = rep(NA,length(colNames)) #setup data, will cause errors if I miss something
    df[day,]$runName = runName
    df[day,]$s=  length(subset(dayState, dayState == 's'))
    df[day,]$i = length(subset(dayState, dayState == 'i'))
    df[day,]$r = length(subset(dayState, dayState == 'r'))
    df[day,]$t = length(subset(dayState, dayState == 't'))
    df[day,]$d_today = length(subset(dayState, dayState == 'd_today'))
    df[day,]$d = length(subset(dayState, dayState == 'd'))
    df[day,]$day = day
    df[day,]$tweets = sum(tweets)
    if (print) {
      cat(
        sprintf(
          "Day=%d, susceptible=%d, infected=%d, recovered=%d, terminal=%d, 
          dead today=%d, dead=%d, tweets=%d, R0=%.2f\n",
          df[day,]$day, df[day,]$s, df[day,]$i, df[day,]$r, df[day,]$t, df[day,]$d_today,
          df[day,]$d, df[day,]$tweets, betaDailyInfectionRates[day]/gammaResolvedPerDayRate))
    }
    tweets = rep(0, nPop) #start fresh every day, certainly wrong.
    for (per in 1:nPop) {
      #end infectious period
      if (dayState[per] == 'i') {
        tweets[per] = rbinom(1, 1, tweetRateInfected)
        if (rbinom(n = 1, size = 1, prob = gammaResolvedPerDayRate) == 1) {
          if (rbinom(n = 1, size = 1, prob = deathRate) == 1) {
            nextDayState[per] = 't' 
          }
          else {
            nextDayState[per] = 'r'
          }
        }
      }
      if (dayState[per] == 't') {
        if (rbinom(n = 1, size =1, prob = 1/meanDaysToDeathFromT) == 1) {
          nextDayState[per] = 'd'
        }
      }
    }
    for (per in 1:nPop) {
      if (dayState[per] == 'i') {
        for (otherPer in sample(1:nPop, nDailyContacts)) {
          if (dayState[otherPer] == 's' &&
              rbinom(n = 1, size = 1, 
                     prob = min(betaDailyInfectionRates[day]/nDailyContacts,1.0)) == 1) {
            nextDayState[otherPer] = 'i'
          }
        }
      }
    }
    dayState = nextDayState
  }
  return(df)
}


nPop = 10000
nWeeks = 10
nDays = nWeeks * 7
gammaResolvedPerDayRateSim = 1/7
tweetRateInfected = .5
deathRateSim = 0.1
meanDaysToDeathFromT = 10

sdOfBetas = .1 # for time varying model of infectiousness
betaInfectionRateMeanSim = .3
betaForWeek = abs(rnorm(nWeeks, betaInfectionRateMeanSim, .2))
betaDailyInfectionRatesSim = rep(betaForWeek, times = rep(7,nWeeks))
nPatientZero = 10
nDailyContacts = 10


simDf = SirtdVaryBeta(runName = 'test', nPop = nPop, nDays = nDays, print = TRUE,
                      betaDailyInfectionRates = betaDailyInfectionRatesSim, 
                      gammaResolvedPerDayRate = gammaResolvedPerDayRateSim,
                      tweetRateInfected = tweetRateInfected, 
                      meanDaysToDeathFromT = meanDaysToDeathFromT,
                      nPatientZero = nPatientZero, 
                      nDailyContacts = nDailyContacts,
                      deathRate = deathRateSim)

```



## Graph of output

```{r}
library(tidyverse)
for (i in 1:nDays) {
  modelPop = sum(simDf[i, c(-1, -2, -9)])
  if (modelPop != nPop) {
    print(sprintf("Non-constant populatin for model, got %d, expected %d"),
          modelPop, nPop)
  }
}
colnames = colnames(simDf)
simLongDf = gather(data = simDf, key = "compartmentSim", value = "count",
                   all_of(colnames[-1:-2]))

plot = ggplot(data = NULL, aes(x = day, y = mean)) +
  geom_point(data = simLongDf, aes(y = count, color = compartmentSim), size = .5) +
  labs(y = "sim data in dots",
       x = "sim days", 
       caption = paste0("dots for simulated truth, compartment is = ") +
  theme(plot.caption=element_text(size=12, hjust=0, margin=margin(15,0,0,0))))

print(plot)

```
